---
title: "Gene Annotation of Harmonised MR Data (PD)"
author: Alexis Felipe Rojas Cruz
output:
  html_document:
    toc: true
    toc_float: true
    theme: cerulean  
    highlight: tango
---

# 1) Setting

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.keep = 'all',
  dpi = 600,
  fig.width = 8,
  fig.height = 8,
  fig.pos = "center"
)
```

# - Packages installation

In this section, we ensure that all necessary packages are installed.

```{r}
# Ensure BiocManager is available
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")

# Required packages
required_packages <- c("tidyverse", "biomaRt", "TwoSampleMR", "readr", "remotes", "devtools")

# Install missing packages
missing_packages <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]
if (length(missing_packages)) BiocManager::install(missing_packages, ask = FALSE)
```

```{r}
# Install TwoSampleMR from GitHub
remotes::install_github("MRCIEU/TwoSampleMR")
# Install MR-PRESSO from Github
devtools::install_github("rondolab/MR-PRESSO")
# Install foresploter from Github
devtools::install_github("adayim/forestploter")
```

# - Loading packages

Here, we load the necessary packages for our analysis.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(biomaRt)
library(TwoSampleMR)
library(MRPRESSO)
library(forestploter)
library(readr)
library(dplyr)
library(grid)
```

# - 2) Load harmonised data

```{r message=FALSE, warning=FALSE}
# Load harmonised data using read.table
pd_data_raw <- read.table("PD_harmonised.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)

# View first few rows
head(pd_data_raw)
```

# - 3) Query Ensembl with biomaRt

```{r message=FALSE, warning=FALSE}
# Extract unique SNPs
snp_list <- unique(pd_data_raw$SNP)

# Use Ensembl SNP mart
ensembl <- useEnsembl(biomart = "ENSEMBL_MART_SNP", dataset = "hsapiens_snp", host = "https://www.ensembl.org")


# Fetch SNP annotations
gene_info <- getBM(
  attributes = c(
    "refsnp_id",
    "chr_name",
    "chrom_start",
    "allele",
    "consequence_type_tv",
    "ensembl_gene_name"
  ),
  filters = "snp_filter",
  values = snp_list,
  mart = ensembl
)

# Get the unique Ensembl gene IDs from SNPs
gene_ids <- unique(gene_info$ensembl_gene_name)

mart <- useMart(
  biomart = "ENSEMBL_MART_ENSEMBL",
  dataset = "hsapiens_gene_ensembl",
  path = "/biomart/martservice",
  host = "https://www.ensembl.org"
)

# Retrieve gene annotations
gene_name_map <- getBM(
  attributes = c(
    "ensembl_gene_id",
    "hgnc_symbol",
    "gene_biotype"
  ),
  filters = "ensembl_gene_id",
  values = gene_ids,
  mart = mart
)


# Merge annotations with harmonised dataset
pd_data <- pd_data_raw %>%
  left_join(gene_info, by = c("SNP" = "refsnp_id")) %>%
  rename(ensembl_gene_id = ensembl_gene_name) %>%
  left_join(gene_name_map, by = "ensembl_gene_id")

# Preview annotated data
head(pd_data)

#save data

write.table(pd_data, file = "PD_data.tsv", sep = "\t", row.names = FALSE, quote = FALSE)
```

# - 4) Perform MR using IVW, WME, MR-Egger

```{r message=FALSE, warning=FALSE}
pd_data <- read.table("PD_data-annotated.tsv", header = TRUE, sep = "\t", stringsAsFactors = FALSE)

PD_MR <- mr(pd_data, method_list = c("mr_ivw", "mr_weighted_median", "mr_wald_ratio", "mr_weighted_mode"))

# Calculate odds ratios
PD_MR <- generate_odds_ratios(PD_MR)

# View MR results
print(PD_MR)

write.csv(PD_MR, "./MR_results/PD_MR_results.csv", row.names = FALSE, quote = FALSE)
```

# - 5) Sensitivity Analyses

# Heterogeneity Test

```{r message=FALSE, warning=FALSE}
heterogeneity <- mr_heterogeneity(pd_data)
print(heterogeneity)

write.csv(heterogeneity, "./Sensitivity_analysis/Heterogeneity/heterogeneity_results.csv", row.names = FALSE, quote = FALSE)

```

# Pleiotropy Test

```{r message=FALSE, warning=FALSE}
pleiotropy <- mr_pleiotropy_test(pd_data)
print(pleiotropy)

write.csv(pleiotropy, "./Sensitivity_analysis/Pleiotropy//pleiotropy_results.csv", row.names = FALSE, quote = FALSE)

```

# MR-PRESSO

```{r message=FALSE, warning=FALSE}
mr_presso <- mr_presso(BetaOutcome = "beta.outcome",
                               BetaExposure = "beta.exposure",
                               SdOutcome = "se.outcome",
                               SdExposure = "se.exposure",
                               OUTLIERtest = TRUE,
                               DISTORTIONtest = TRUE,
                               data = pd_data,
                               NbDistribution = 1000,
                               SignifThreshold = 0.05)
print(mr_presso)

write.csv(mr_presso, "./Sensitivity_analysis/MR-PRESSO/mr-presso_results.csv", row.names = FALSE, quote = FALSE)

```

# Leave-One-Out Analysis

```{r message=FALSE, warning=FALSE}
loo <- mr_leaveoneout(pd_data)
print(loo)

write.csv(loo, "./Sensitivity_analysis/Leave-one-out/leave_one_out_results.csv", row.names = FALSE, quote = FALSE)

```

# - 6 Plots

# MR Scatter Plot

```{r message=FALSE, warning=FALSE}
scatter_plot <- mr_scatter_plot(PD_MR, pd_data)
scatter_plot
ggsave("./MR_results/MR_scatter_plot.pdf", scatter_plot[[1]], width = 7, height = 5)
```

# Leave-One-Out Plot

```{r message=FALSE, warning=FALSE}
loo_plot <- mr_leaveoneout_plot(loo)
loo_plot
ggsave("./Sensitivity_analysis/Leave-one-out/leave_one_out_plot.pdf", loo_plot[[1]], width = 7, height = 5)
```

# Forest Plot

```{r message=FALSE, warning=FALSE}
forest_plot <- mr_forest_plot(loo)
forest_plot
ggsave("./MR_results/forest_plot.pdf", forest_plot[[1]], width = 7, height = 8)
```

# Funnel Plot

```{r message=FALSE, warning=FALSE}
PD_MR_singlesnp <- mr_singlesnp(pd_data)

funnel_plot <- mr_funnel_plot(PD_MR_singlesnp)
funnel_plot
ggsave("./MR_results/funnel_plot.pdf", funnel_plot[[1]], width = 7, height = 8)
```
# Forest plot

```{r message=FALSE, warning=FALSE}

# DATA PREPARATION 

# Format OR and 95% CI 
PD_MR$`OR (95% CI)` <- sprintf("%.2f (%.2f to %.2f)",
                                       PD_MR$or,
                                       PD_MR$or_lci95,
                                       PD_MR$or_uci95)

# Format p-values: scientific notation if <0.001, else 3 decimal places
PD_MR$`P-value` <- ifelse(PD_MR$pval < 0.001,
                                  formatC(PD_MR$pval, format = "e", digits = 2),
                                  sprintf("%.3f", PD_MR$pval))

# Calculate standard error from confidence intervals
PD_MR$se <- (log(PD_MR$or_uci95) - log(PD_MR$or))/1.96

# Create blank column that will hold the confidence interval graphics
PD_MR$` ` <- paste(rep(" ", 25), collapse = " ")  

# CREATE PLOT DATA FRAME 

# Assemble all components into a data frame for plotting
dt <- data.frame(
  Method = PD_MR$method,      
  SNPs = as.character(PD_MR$nsnp),  
  `OR (95% CI)` = PD_MR$`OR (95% CI)`,  
  ` ` = PD_MR$` `,            
  `P-value` = PD_MR$`P-value`, 
  
  # Numeric versions for calculations (not displayed directly)
  estimate = PD_MR$or,        
  low = PD_MR$or_lci95,      
  high = PD_MR$or_uci95,      
  se = PD_MR$se,            
  
  stringsAsFactors = FALSE,
  check.names = FALSE  # Preserve special characters in column names
)

# DEFINE PLOT THEME 

# Customize all visual aspects of the forest plot
tm <- forest_theme(
  # Text sizing
  base_size = 12,             
  base_family = "Helvetica",       
  
  # Confidence interval appearance
  ci_pch = 18,                
  ci_col = "#1E64B0",         
  ci_lwd = 2,                
  ci_alpha = 0.8,             
  ci_Theight = 0.3,           
  
  # Reference line at OR=1
  refline_col = "#E41A1C",    
  refline_lwd = 1.5,         
  refline_lty = "dashed", 
  
  # Spacing and layout
  colgap = unit(8, "mm"),    
  header_height = unit(20, "mm"), 
  plot_margin = unit(c(1, 2, 1, 1), "cm")  
)

# CREATE FOREST PLOT 

# Generate the forest plot object
p <- forest(
  # Columns to display (in order)
  dt[, c("Method", "SNPs", " ", "OR (95% CI)", "P-value")],
  
 
  est = dt$estimate,   
  lower = dt$low,      
  upper = dt$high,    
  sizes = 0.4,         
  
  # Plot configuration
  ci_column = 3,      
  ref_line = 1,        
  arrow_lab = c("Protective", "Risk"),  
  x_trans = "log",     
  xlim = c(0.3, 3),    
  ticks_at = c(0.3, 0.5, 1, 2, 3),  
  
  # Apply our custom theme
  theme = tm
)

# DISPLAY THE PLOT 

# Render the plot with specified dimensions
plot(p, vp = viewport(width = unit(8, "inches"), 
                     height = unit(2, "inches")))

# To save: 
pdf("./Forest/forest_plot.pdf", width = 8, height = 2)  # Adjust dimensions as needed
plot(p)  # Your forest plot object
dev.off()  # Close the graphics device

```

# Forest plot for each SNP

```{r message=FALSE, warning=FALSE}

pd_snps <- pd_data %>%
  mutate(
    or = exp(beta.outcome),
    or_lci95 = exp(beta.outcome - 1.96 * se.outcome),
    or_uci95 = exp(beta.outcome + 1.96 * se.outcome)
  ) %>%
  dplyr::select(
    Region,
    taxon,
    SNP,
    hgnc_symbol,
    pval.outcome,
    or,
    or_lci95,
    or_uci95
  )
print(pd_snps)
```


```{r message=FALSE, warning=FALSE}

# DATA PREPARATION

# SORT BY P-VALUE (ascending order)
pd_snps <- pd_snps[order(pd_snps$pval.outcome), ]

# Convert SNP names to factor to preserve sorting order
pd_snps$SNP <- factor(pd_snps$SNP, levels = pd_snps$SNP)

# Format OR and 95% CI 
pd_snps$`OR (95% CI)` <- sprintf("%.2f (%.2f to %.2f)",
                               pd_snps$or,
                               pd_snps$or_lci95,
                               pd_snps$or_uci95)

# Format p-values
pd_snps$`P-value` <- ifelse(pd_snps$pval.outcome < 0.001,
                          formatC(pd_snps$pval.outcome, format = "e", digits = 2),
                          sprintf("%.3f", pd_snps$pval.outcome))

# Create blank column for forest plot visualization (immediately after Gene)
pd_snps$` ` <- paste(rep(" ", 20), collapse = " ")  

# CREATE PLOT DATA FRAME
dt <- data.frame(
  Region = pd_snps$Region,
  Taxon = pd_snps$taxon,
  SNP = pd_snps$SNP,
  Gene = pd_snps$hgnc_symbol,
  ` ` = pd_snps$` `,            # Forest plot visualization goes here
  `OR (95% CI)` = pd_snps$`OR (95% CI)`,  
  `P-value` = pd_snps$`P-value`,
  
  # Numeric versions for plotting
  estimate = pd_snps$or,
  low = pd_snps$or_lci95,
  high = pd_snps$or_uci95,
  stringsAsFactors = FALSE,
  check.names = FALSE
)

# DEFINE PLOT THEME
tm <- forest_theme(
  base_size = 10,
  base_family = "Helvetica",
  ci_pch = 18,
  ci_col = "#1E64B0",
  ci_lwd = 2,
  ci_alpha = 0.8,
  ci_Theight = 0.3,
  refline_col = "#E41A1C",
  refline_lwd = 1.5,
  refline_lty = "dashed",
  colgap = unit(6, "mm"),
  header_height = unit(25, "mm"),
  plot_margin = unit(c(1, 2, 1, 1), "cm")
)

# CREATE FOREST PLOT
p2 <- forest(
  # Column order with visualization after Gene
  dt[, c("Region", "Taxon", "SNP", "Gene", " ", "OR (95% CI)", "P-value")],
  
  est = dt$estimate,
  lower = dt$low,
  upper = dt$high,
  sizes = 0.4,
  ci_column = 5,  # This is now the blank column after Gene
  ref_line = 1,
  arrow_lab = c("Protective", "Risk"),
  x_trans = "log",
  xlim = c(0.8, 1.2),
  ticks_at = c(0.8, 0.9, 1.0, 1.1, 1.2),
  theme = tm
)

# DISPLAY THE PLOT
plot(p2, vp = viewport(width = unit(8, "inches"), 
                     height = unit(2.5, "inches")))

# To save:
pdf("./Forest/pd_snps_forest_plot.pdf", width = 8, height = 2.5)
plot(p2)
dev.off()

```

# Session information

```{r, session-information}
sessionInfo()
```

