---
title: "Metagenomics and Metabolomics Integration"
author: "Alexis Felipe Rojas Cruz"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cerulean  
    highlight: tango
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.keep = 'all'
)
```

# 1. Install and Load Required Packages

```{r include=FALSE, comment = ""}
# Define packages and their sources
bioc_packages <- c("ComplexHeatmap", "clusterProfiler", "DOSE", "enrichplot", "AnnotationDbi", "GO.db", "KEGGREST")
cran_packages <- c("RColorBrewer", "tidyverse", "ggplot2", "ggrepel", "pheatmap", "readr", "ggvenn", "circlize", "networkD3", "webshot2", "htmlwidgets", "patchwork")

# Install Bioconductor packages
missing_bioc <- bioc_packages[!bioc_packages %in% rownames(installed.packages())]
if (length(missing_bioc) > 0) {
  if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
  }
  BiocManager::install(missing_bioc)
}

# Install CRAN packages
missing_cran <- cran_packages[!cran_packages %in% rownames(installed.packages())]
if (length(missing_cran) > 0) {
  install.packages(missing_cran)
}

# Load all libraries
library(ComplexHeatmap)
library(clusterProfiler)
library(DOSE)
library(enrichplot)
library(AnnotationDbi)
library(KEGGREST)
library(tidyverse)
library(pheatmap)
library(ggrepel)
library(RColorBrewer)
library(readr)
library(ggvenn)
library(circlize)
library(networkD3)
library(webshot2)
library(htmlwidgets)
library(patchwork)
library(dplyr)
```

## - Loading palettes

In this section, we load additional color palettes and define some custom ones.

```{r message=FALSE, warning=FALSE}
# Load additional colour palette packages
library(unikn, quietly = TRUE)
library(RColorBrewer, quietly = TRUE)
library(yarrr, quietly = TRUE)
library(scales, quietly = TRUE)
library(ggsci, quietly = TRUE)

# Define a set of custom color palettes from the unikn package
mix_1 <- usecol(pal = c(Karpfenblau, "white", Peach), n = 15)
mix_2 <- usecol(pal = c(rev(pal_seeblau), "white", pal_pinky))
mix_3 <- usecol(pal = c(rev(pal_bordeaux), "white", pal_petrol), n = 15)

# Display the custom palettes
seecol(list(mix_1, mix_2, mix_3), col_brd = "white", lwd_brd = 4, title = "Comparing palettes mixed from unikn colors", pal_names = c("mix_1", "mix_2", "mix_3"))

# Define a second set of custom palettes from the RColorBrewer and yarrr packages
brew_mix <- usecol(c(rev(brewer.pal(n = 4, name = "Reds")), "white", brewer.pal(n = 4, name = "Blues")), n = 13)
brew_ext <- usecol(brewer.pal(n = 11, name = "Spectral"), n = 12)
yarrr_mix <- usecol(c(piratepal("nemo"), piratepal("bugs")))
yarrr_mod <- usecol(c(piratepal("ipod")), n = 9)

# Display the second set of custom palettes
seecol(pal = list(brew_mix, brew_ext, yarrr_mix, yarrr_mod), col_brd = "white", lwd_brd = 2, title = "Using usecol() and seecol() to mix and modify palettes", pal_names = c("brew_mix", "brew_ext", "yarrr_mix", "yarrr_mod"))

# Define additional custom palettes from the scales package
natjournals_palette <- pal_npg("nrc")(10)
```

## Loading data


```{r load-data}
# EUR CLR matrix
EUR_clr <- readRDS("./Data/EUR_clr_counts.rds")

# Metabolites (log10 transformed)
metabolites_raw <- read.csv("./Data/Differential_metabolites.csv",
                              header = TRUE,
                              row.names =1,
                              sep = ";",
                              check.names = FALSE)

# MAG based on genus
MAG_genus <- read.csv("./Data/MAGs_genus.csv",
                              header = TRUE,
                              check.names = FALSE)

```

## Preparing data

```{r prepare-data}
# Filter EUR_clr
EUR_clr_filt <- EUR_clr[rownames(EUR_clr) %in% MAG_genus$taxon, ] %>%
  rename_with(~ sub("^E\\.", "", .x), starts_with("E.")) %>%
  as.matrix()

# Set up metabolites
metabolites <- metabolites_raw %>%
  as.data.frame() %>%
  dplyr::select(dplyr::matches("^(CT_|PD_)")) %>%
  as.matrix()

# Identify common sample columns
common_samples <- intersect(
  colnames(EUR_clr_filt),
  colnames(metabolites)
)

# Reorder both matrices
EUR_clr_filt <- EUR_clr_filt[, common_samples, drop = FALSE]
metabolites <- metabolites[, common_samples, drop = FALSE]

identical(
  colnames(EUR_clr_filt),
  colnames(metabolites)
)

cat("EUR_clr dimensions:", dim(EUR_clr_filt), "\n")
cat("Metabolites dimensions:", dim(metabolites), "\n")
cat("Number of samples:", length(common_samples), "\n")

```
## Define group vector

```{r group vector}
group_vec <- ifelse(
  grepl("^PD_", colnames(EUR_clr_filt)), "PD",
  ifelse(grepl("^CT_", colnames(EUR_clr_filt)), "Healthy", NA)
)

names(group_vec) <- colnames(EUR_clr_filt)

group_vec <- factor(group_vec, levels = c("PD", "Healthy"))

PD_samples <- names(group_vec)[group_vec == "PD"]
H_samples  <- names(group_vec)[group_vec == "Healthy"]

# check
all(PD_samples %in% colnames(EUR_clr_filt))
all(H_samples  %in% colnames(EUR_clr_filt))

```


# 1. ASV–Metabolite Spearman correlations

```{r Spearman and Permutation functions}
# =====================================================
# Spearman correlation function
# =====================================================
S_correlation <- function(x, y, min_n = 3) {

  ok <- complete.cases(x, y)

  if (sum(ok) < min_n ||
      sd(x[ok]) == 0 ||
      sd(y[ok]) == 0) {
    return(c(r = NA_real_, p = NA_real_))
  }

  test <- suppressWarnings(
    cor.test(x[ok], y[ok], method = "spearman", exact = FALSE)
  )

  c(
    r = unname(test$estimate),
    p = test$p.value
  )
}


# =====================================================
# Permutation test for Δρ
# =====================================================
perm_delta_rho <- function(x, y, group,
                           n_perm = 5000,
                           min_n = 3) {

  # Remove missing values
  ok <- complete.cases(x, y, group)
  x <- x[ok]
  y <- y[ok]
  g <- factor(group[ok])

  groups <- levels(g)

  # Only two groups allowed
  if (length(groups) != 2) {
    return(c(delta_rho = NA_real_,
             p_perm   = NA_real_,
             n_eff    = NA_real_))
  }

  g1 <- groups[1]
  g2 <- groups[2]

  n1 <- sum(g == g1)
  n2 <- sum(g == g2)

  # Minimum group size
  if (n1 < min_n || n2 < min_n) {
    return(c(delta_rho = NA_real_,
             p_perm   = NA_real_,
             n_eff    = min(n1, n2)))
  }

  # Zero-variance checks
  if (sd(x[g == g1]) == 0 || sd(y[g == g1]) == 0 ||
      sd(x[g == g2]) == 0 || sd(y[g == g2]) == 0) {
    return(c(delta_rho = NA_real_,
             p_perm   = NA_real_,
             n_eff    = min(n1, n2)))
  }

  # Safe Spearman
  safe_spearman <- function(x_sub, y_sub) {
    if (length(x_sub) < 2 ||
        sd(x_sub) == 0 ||
        sd(y_sub) == 0) {
      return(NA_real_)
    }
    tryCatch(
      cor(x_sub, y_sub,
          method = "spearman",
          use = "complete.obs"),
      error = function(e) NA_real_
    )
  }

  # Observed correlations
  rho1 <- safe_spearman(x[g == g1], y[g == g1])
  rho2 <- safe_spearman(x[g == g2], y[g == g2])

  if (is.na(rho1) || is.na(rho2)) {
    return(c(delta_rho = NA_real_,
             p_perm   = NA_real_,
             n_eff    = min(n1, n2)))
  }

  obs_delta <- rho1 - rho2

  # Permutations
  perm_deltas <- numeric(n_perm)
  n_total <- length(g)

  for (i in seq_len(n_perm)) {

    perm_g1 <- sample(n_total, n1)
    perm_g2 <- setdiff(seq_len(n_total), perm_g1)

    rho1_p <- safe_spearman(x[perm_g1], y[perm_g1])
    rho2_p <- safe_spearman(x[perm_g2], y[perm_g2])

    perm_deltas[i] <-
      ifelse(!is.na(rho1_p) && !is.na(rho2_p),
             rho1_p - rho2_p,
             NA_real_)
  }

  valid_perms <- sum(!is.na(perm_deltas))

  if (valid_perms == 0) {
    return(c(delta_rho = obs_delta,
             p_perm   = NA_real_,
             n_eff    = min(n1, n2)))
  }

  p_perm <- (sum(abs(perm_deltas) >= abs(obs_delta), na.rm = TRUE) + 1) /
            (valid_perms + 1)

  c(
    delta_rho = obs_delta,
    p_perm   = p_perm,
    n_eff    = min(n1, n2),
    rho_PD   = rho1,
    rho_H    = rho2
  )
}

```

## ASV–Metabolite correlations

```{r ASV-Metabolite correlations}

# ASV–Metabolite correlations

ASV_meta_cor <- expand_grid(
  ASV        = rownames(EUR_clr_filt),
  Metabolite = rownames(metabolites)
) %>%
  mutate(

    # PD-only correlations
    PD_stats = map2(ASV, Metabolite, ~
      S_correlation(
        EUR_clr_filt[.x, PD_samples, drop = TRUE],
        metabolites[.y, PD_samples, drop = TRUE]
      )
    ),

    # Healthy-only correlations
    H_stats = map2(ASV, Metabolite, ~
      S_correlation(
        EUR_clr_filt[.x, H_samples, drop = TRUE],
        metabolites[.y, H_samples, drop = TRUE]
      )
    )
  ) %>%
  mutate(
    rho_PD    = map_dbl(PD_stats, "r"),
    p_PD      = map_dbl(PD_stats, "p"),
    rho_H     = map_dbl(H_stats, "r"),
    p_H       = map_dbl(H_stats, "p"),
    delta_rho = rho_PD - rho_H
  ) %>%
  dplyr::select(-PD_stats, -H_stats)

# =====================================================
# Permutation Δρ
# =====================================================
ASV_meta_cor <- ASV_meta_cor %>%
  mutate(
    perm_stats = map2(ASV, Metabolite, ~
      perm_delta_rho(
        x     = EUR_clr_filt[.x, ],
        y     = metabolites[.y, ],
        group = group_vec,
        n_perm = 5000,
        min_n  = 3
      )
    ),
    delta_rho_perm = map_dbl(perm_stats, "delta_rho"),
    p_delta        = map_dbl(perm_stats, "p_perm")
  ) %>%
  dplyr::select(-perm_stats)
```

## Multiple testing correction

```{r testing correction}
ASV_meta_cor <- ASV_meta_cor %>%
  mutate(
    p_adj_PD    = p.adjust(p_PD,    method = "BH"),
    p_adj_H     = p.adjust(p_H,     method = "BH"),
    p_adj_delta = p.adjust(p_delta, method = "BH")
  ) %>%
  arrange(desc(abs(delta_rho_perm))) %>%
  mutate(
    delta_category = case_when(
      abs(delta_rho_perm) >= 0.5 ~ "Divergent",
      abs(delta_rho_perm) >= 0.2 ~ "Moderately concordant",
      TRUE                       ~ "Highly concordant"
    ),
    direction_rho = case_when(
      delta_rho_perm > 0 ~ "Healthy > PD",
      delta_rho_perm < 0 ~ "PD > Healthy",
      TRUE ~ NA_character_
    )
  )


```

## Data organization 

```{r data-organization}

# ===================================================== 
# 1. Build combined effect-size matrix
# =====================================================
heatmap_cor <- ASV_meta_cor %>%
  transmute(
    Pair = paste(ASV, Metabolite, sep = " | "),
    PD      = rho_PD,
    Healthy = rho_H
  ) %>%
  column_to_rownames("Pair") %>%
  as.matrix()

# =====================================================
# 2. Build combined p-value matrix
# =====================================================
heatmap_pvals <- ASV_meta_cor %>%
  transmute(
    Pair = paste(ASV, Metabolite, sep = " | "),
    PD     = p_adj_PD,
    Healthy = p_adj_H,
  ) %>%
  column_to_rownames("Pair") %>%
  as.matrix()


# =====================================================
# 3. Δρ (difference in Spearman correlations)
# =====================================================
ASV_meta_cor$delta_category <- factor(
  ASV_meta_cor$delta_category,
  levels = c(
    "Highly concordant",
    "Moderately concordant",
    "Divergent"
  )
)

delta_category <- factor(
  ASV_meta_cor$delta_category,
  levels = c("Highly concordant", "Moderately concordant", "Divergent")
)

```


```{r heatmap}

# =====================================================
# 1. Color palettes (data-driven)
# =====================================================
rho_lim <- max(abs(heatmap_cor), na.rm = TRUE)
breaks_val <- seq(-rho_lim, rho_lim, length.out = 13)

nature_slate_rust <- colorRampPalette(
  c("#4A6FA5", "white", "#B5543A"))(13)

col_heatmap_guide <- colorRamp2(
  breaks = breaks_val,
  colors = nature_slate_rust
)

# Column group colors for PD vs Healthy
group_colors <- c(
  "PD"      = "#8B1E3F",  # muted burgundy
  "Healthy" = "#355F8C"   # muted steel blue
)

# Row annotation colors based on category
delta_colors <- c(
  "Highly concordant"     = "#4E7F5D",  # muted green
  "Moderately concordant" = "#C2A14D",  # muted ochre
  "Divergent"             = "#8C4A2F"   # muted brown-red
)

# =====================================================
# 2. Column annotation
# =====================================================
col_anno <- HeatmapAnnotation(
  Group = factor(colnames(heatmap_cor), levels = colnames(heatmap_cor)),
  col = list(Group = group_colors),
  annotation_name_side = "left",
  simple_anno_size = unit(4, "mm"),
  annotation_legend_param = NULL
)

# =====================================================
# 3. Row annotation using category
# =====================================================
row_anno <- rowAnnotation(
  Category = delta_category,
  col = list(Category = delta_colors),
  annotation_width = unit(4, "mm"),
  show_annotation_name = FALSE,
  annotation_legend_param = list(
    title = "Δp category",
    at = c("Highly concordant", "Moderately concordant", "Divergent"),
    legend_direction = "vertical"
  )
)

# =====================================================
# 4. Build Heatmap
# =====================================================
Heatmap_combined <- Heatmap(
  heatmap_cor,
  top_annotation  = col_anno,
  left_annotation = row_anno,
  col             = col_heatmap_guide,

  cell_fun = function(j, i, x, y, w, h, fill) {
    p_val <- heatmap_pvals[i, j]
    if (!is.na(p_val)) {
      if (p_val < 0.01) {
        grid.text("**", x, y, gp = gpar(fontsize = 15))
      } else if (p_val < 0.05) {
        grid.text("*", x, y, gp = gpar(fontsize = 15))
      }
    }
  },

  row_split = ASV_meta_cor$delta_category,
  row_title = NULL,
  cluster_row_slices = FALSE,
  cluster_rows = FALSE,          

  use_raster      = TRUE,
  raster_quality  = 3,
  cluster_columns = FALSE,
  border          = TRUE,
  rect_gp         = gpar(col = NA),
  show_row_dend   = FALSE,
  show_row_names  = FALSE,
  row_names_gp    = gpar(fontsize = 7),
  column_names_gp = gpar(fontsize = 10, fontface = "bold"),

  heatmap_legend_param = list(
    title = expression("Spearman "~rho),
    legend_direction = "vertical",
    at = round(seq(-rho_lim, rho_lim, length.out = 5), 2)
  )
)

# =====================================================
# 5. Draw Heatmap
# =====================================================
draw(
  Heatmap_combined,
  heatmap_legend_side      = "right",
  annotation_legend_side   = "right",
  merge_legends            = TRUE,
  align_annotation_legend  = TRUE,
  newpage                  = TRUE
)

# Capture the plot
heatmap_plot <- recordPlot()
```

```{r save}
# =====================================================
# Save heatmap as PDF
# =====================================================
pdf("Spearman/heatmap_plot.pdf", width = 10, height = 18)
replayPlot(heatmap_plot)
dev.off()
# Open a new PDF device (portrait layout)
pdf(file = "./Spearman/Heatmap.pdf", width = 10, height = 14)

# Draw your heatmap object
draw(
  Heatmap_combined,
  heatmap_legend_side = "right",
  annotation_legend_side = "right",
  merge_legends = TRUE,                 
  align_annotation_legend = TRUE,      
  newpage = TRUE                        
)

# Close the PDF device
dev.off()

```

```{r save_cor_df}

# Save the correlation table to CSV
write.csv(
  ASV_meta_cor,
  file = "./Spearman/asv_metabolites.csv",  
  row.names = FALSE
)

# Quick confirmation
cat("Correlation table saved\n")
```

# 2. KEGG annotation: KO → Reaction → Metabolite

```{r load-data}
# Correlation data
ASV_meta_cor <- read.csv("./Spearman/asv_metabolites.csv", header = TRUE, stringsAsFactors = FALSE)

# EggNOG annotations
eggnog <- read.delim("./Data/Final_annotation.tsv", sep = "\t", header = TRUE)

```

## Clean eggnog

```{r eggnog clean}
# Keep only eggNOG genes present in the MAG_genus dataset

eggnog_clean <- eggnog %>%
  filter(
    !is.na(KEGG_ko), KEGG_ko != "-",
    !is.na(KEGG_Module), KEGG_Module != "-",
    !is.na(KEGG_Pathway), KEGG_Pathway != "-"
  ) %>%
  separate_rows(GOs, sep = "[,;]") %>%
  separate_rows(KEGG_ko, sep = "[,;]") %>%
  separate_rows(KEGG_Module, sep = "[,;]") %>%
  separate_rows(KEGG_Pathway, sep = "[,;]") %>%
  mutate(
    KEGG_ko     = stringr::str_trim(KEGG_ko),
    KEGG_Module = stringr::str_trim(KEGG_Module),
    KEGG_Pathway = stringr::str_trim(KEGG_Pathway)
  ) %>%
  select(
    gene_id = query,
    MAG_ID  = MAG_gene,
    KEGG_ko,
    KEGG_Module,
    KEGG_Pathway
  ) %>%
  filter(grepl("^ko", KEGG_Pathway)) %>%
  inner_join(MAG_genus %>% select(gene_id,
                                  Group_ASV = Group), by = "gene_id") %>%
  distinct()
```

## KEGG functions

```{r kegg_functions, message=FALSE, warning=FALSE}

# =====================================================
# 1. KO → Reaction
# =====================================================
get_ko_reactions <- function(ko_vec) {

  ko_vec %>%
    unique() %>%
    na.omit() %>%
    map_dfr(function(ko_raw) {

      ko_id <- sub("^ko:", "", ko_raw)

      rxn <- tryCatch(
        keggLink("reaction", paste0("ko:", ko_id)),
        error = function(e) NULL
      )

      if (is.null(rxn) || length(rxn) == 0) {
        tibble(
          KEGG_ko = paste0("ko:", ko_id),
          Reaction = NA_character_
        )
      } else {
        tibble(
          KEGG_ko = paste0("ko:", ko_id),
          Reaction = sub("rn:", "", rxn)
        )
      }
    })
}

ko_rxn <- get_ko_reactions(eggnog_clean$KEGG_ko)

# =====================================================
# 2. Reaction → Equation
# =====================================================
get_reaction_equations <- function(ko_rxn_table) {

  ko_rxn_table %>%
    select(Reaction) %>%
    distinct() %>%
    pull(Reaction) %>%
    na.omit() %>%
    map_dfr(function(rxn_raw) {

      rxn_id <- sub("^rn:", "", rxn_raw)

      info <- tryCatch(
        keggGet(paste0("rn:", rxn_id))[[1]],
        error = function(e) NULL
      )

      if (is.null(info) || is.null(info$EQUATION)) {
        tibble(
          Reaction = rxn_id,
          Equation = NA_character_
        )
      } else {
        tibble(
          Reaction = rxn_id,
          Equation = info$EQUATION
        )
      }
    })
}

rxn_eq <- get_reaction_equations(ko_rxn)

# =====================================================
# 3. Reaction → Substrates / Products
# =====================================================
parse_reaction_compounds <- function(rxn_eq_table) {

  rxn_eq_table %>%
    filter(!is.na(Equation)) %>%
    select(Reaction, Equation) %>%
    pmap_dfr(function(Reaction, Equation) {

      sides <- strsplit(Equation, "\\s*<?=>\\s*")[[1]]
      if (length(sides) != 2) return(NULL)

      Substrates <- stringr::str_extract_all(sides[1], "[A-Z]\\d{5}")[[1]]
      Products   <- stringr::str_extract_all(sides[2], "[A-Z]\\d{5}")[[1]]

      bind_rows(
        tibble(
          Reaction = Reaction,
          Compound = Substrates,
          Role = "Substrate"
        ),
        tibble(
          Reaction = Reaction,
          Compound = Products,
          Role = "Product"
        )
      )
    }) %>%
    filter(!is.na(Compound)) %>%
    distinct()
}


rxn_compounds <- parse_reaction_compounds(rxn_eq)


# =====================================================
# 4. Compound → Compound Name
# =====================================================
get_compound_names <- function(rxn_compounds_table) {

  rxn_compounds_table %>%
    select(Compound) %>%
    distinct() %>%
    pull(Compound) %>%
    na.omit() %>%
    map_dfr(function(cid_raw) {

      cid <- sub("^cpd:", "", cid_raw)

      info <- tryCatch(
        keggGet(paste0("cpd:", cid))[[1]],
        error = function(e) NULL
      )

      if (is.null(info) || is.null(info$NAME)) {
        tibble(
          Compound = cid,
          KEGG_Metabolite = NA_character_
        )
      } else {
        tibble(
          Compound = cid,
          KEGG_Metabolite = info$NAME[1]
        )
      }
    })
}

compound_names <- get_compound_names(rxn_compounds)


# =====================================================
# 5. Pathway → Description (annotation only)
# =====================================================
get_pathway_descriptions <- function(pathway_vec) {

  pathway_vec %>%
    unique() %>%
    na.omit() %>%
    map_dfr(function(pid_raw) {

      pid <- sub("^path:", "", pid_raw)

      info <- tryCatch(
        keggGet(pid)[[1]],
        error = function(e) NULL
      )

      if (is.null(info) || is.null(info$NAME)) {
        tibble(
          KEGG_Pathway = pid,
          Pathway_description = NA_character_
        )
      } else {
        tibble(
          KEGG_Pathway = pid,
          Pathway_description = paste(info$NAME, collapse = "; ")
        )
      }
    })
}

pathway_desc <- get_pathway_descriptions(eggnog_clean$KEGG_Pathway)


```

## KEGG annotations

```{r kegg_annotations, message=FALSE, warning=FALSE}

# Integration
kegg_table <- eggnog_clean %>%
  left_join(ko_rxn, by = "KEGG_ko") %>%
  left_join(rxn_compounds, by = "Reaction") %>%
  left_join(compound_names, by = "Compound") %>%
  left_join(pathway_desc, by = "KEGG_Pathway") %>%
  mutate(
    KEGG_Metabolite = KEGG_Metabolite %>%
      stringr::str_trim() %>%
      stringr::str_remove(";+$")
  )

```

## Filtering data based on AVS

```{r KEGG_taxa assignments}

AVS_filtered <- kegg_table %>%
  
  # Join KEGG-annotated genes to MAGs and taxonomic assignments
  inner_join(
    MAG_genus %>% 
      select(gene_id, taxon, ends_with(".MAG"), -classification.MAG),
    by = "gene_id"
  ) %>%
  
  # Integrate ASV–metabolite correlation results at the taxon level
  inner_join(
    ASV_meta_cor %>% 
      select(ASV, Metabolite, delta_rho_perm, p_delta, delta_category),
    relationship = "many-to-many",
    by = c("taxon" = "ASV")
  ) %>%
  
  # Annotate metabolites with group-level metadata
  left_join(
    metabolites_raw %>% 
      rownames_to_column("Metabolite") %>% 
      select(Metabolite, Group_Meta = Group),
    by = "Metabolite"
  ) %>%
  
  # Apply statistical significance and effect-size thresholds
  filter(p_delta < 0.05, abs(delta_rho_perm) >= 0.2) %>%
  
  # Classify biologically consistent versus discordant associations
  mutate(
    Agreement = case_when(
      Group_ASV == "Healthy" & Group_Meta == "Healthy" & delta_rho_perm > 0 ~ "Healthy-consistent",
      Group_ASV == "PD"      & Group_Meta == "PD"      & delta_rho_perm < 0 ~ "PD-consistent",
      TRUE ~ "Discordant"
    )
  ) %>%
  
  # Retain only biologically consistent AVS associations
  filter(Agreement != "Discordant") %>%
  distinct()

```

## PD

```{r sankey-PD}
sankey_PD <- AVS_filtered %>%
  filter(Agreement == "PD-consistent") %>%
  filter(!is.na(Pathway_description)) %>%
  separate(MAG_ID,
           into = c("Prefix", "MAG", "Gene"),
           sep = "__",
           remove = TRUE) %>%
  mutate(MAG = gsub("^[a-z]__", "", MAG)) %>%
  select(MAG, Gene, KEGG_Module, Pathway_description, Metabolite) %>%
  distinct()

# Save to CSV
write.csv(
  sankey_PD,
  file = "./Sankey/PD_sankey_table.csv",
  row.names = FALSE,
  quote = FALSE
)

```


```{r sankey-links}
# Define functional hierarchy
func_levels <- c(
  "MAG",
  "Gene",
  "KEGG_Module",
  "Pathway_description",
  "Metabolite"
)

# Build functional connections
Sankey_links <- sankey_PD %>%
  select(
    MAG,
    Gene,
    KEGG_Module,
    Pathway_description,
    Metabolite
  ) %>%
  distinct() %>%
  mutate(row_id = row_number()) %>%
  pivot_longer(
    cols = all_of(func_levels), 
    names_to = "level",
    values_to = "node"
  ) %>%
  mutate(level = factor(level, levels = func_levels)) %>%
  arrange(row_id, level) %>%
  group_by(row_id) %>%
  mutate(
    next_node = lead(node),
    MAG_link = first(node[level == "MAG"])
  ) %>%
  ungroup() %>%
  filter(!is.na(next_node)) %>%
  count(MAG_link, node, next_node, name = "value") %>%
  rename(
    source = node,
    target = next_node,
    MAG = MAG_link
  )


```

```{r sankey-nodes}
# Build nodes
Sankey_nodes <- data.frame(
  name = unique(c(Sankey_links$source, Sankey_links$target)),
  stringsAsFactors = FALSE
)

# Assign node groups by hierarchy level
Sankey_nodes <- Sankey_nodes %>%
  mutate(Level = case_when(
    name %in% sankey_PD$MAG ~ "MAG",
    name %in% sankey_PD$Gene ~ "Gene",
    name %in% sankey_PD$KEGG_Module ~ "KEGG_Module",
    name %in% sankey_PD$Pathway_description ~ "KEGG_Pathway",
    name %in% sankey_PD$Metabolite ~ "Metabolite",
    TRUE ~ "Other"
  ))

# Arrange nodes by hierarchy order
Sankey_nodes <- Sankey_nodes %>%
  mutate(Level = factor(
    Level,
    levels = c("MAG", "Gene", "KEGG_Module", "KEGG_Pathway", "Metabolite")
  )) %>%
  arrange(Level)

# Add numeric IDs after arranging
Sankey_nodes <- Sankey_nodes %>%
  mutate(nodeID = row_number() - 1)

Sankey_links <- Sankey_links %>%
  mutate(
    IDsource = match(source, Sankey_nodes$name) - 1,
    IDtarget = match(target, Sankey_nodes$name) - 1
  )

```

```{r sankey-plot, echo=FALSE, message=FALSE, warning=FALSE}
# Generate palette
# Node levels
level_colors <- c(
  MAG           = pal_npg("nrc")(10)[1],
  Gene          = pal_npg("nrc")(10)[3],
  KEGG_Module   = pal_npg("nrc")(10)[4],
  KEGG_Pathway  = pal_npg("nrc")(10)[2],
  Metabolite    = pal_npg("nrc")(10)[5]
)

# MAGs for link colors
base_palette <- c(
  "#0072B2",  # blue
  "#D55E00",  # vermillion/red
  "#009E73",  # green
  "#CC79A7",  # pink/magenta
  "#F0E442",  # yellow
  "#56B4E9",  # sky blue
  "#E69F00",  # orange
  "#000000"   # black
)

MAGs <- unique(Sankey_links$MAG)
n_mag <- length(MAGs)
mag_palette <- if (n_mag <= length(base_palette)) {
  base_palette[1:n_mag]
} else {
  colorRampPalette(base_palette)(n_mag)
}

# Prepare JS domain/range strings for injection
mag_domain <- paste0('"', MAGs, '"', collapse = ",")
mag_range  <- paste0('"', mag_palette, '"', collapse = ",")

level_domain <- paste0('"', names(level_colors), '"', collapse = ",")
level_range  <- paste0('"', level_colors, '"', collapse = ",")
```

```{r sankey-plot, echo=FALSE, message=FALSE, warning=FALSE}

# -----------------------------
# 1. Create Sankey network
# -----------------------------
sankey_plot <- sankeyNetwork(
  Links = Sankey_links,
  Nodes = Sankey_nodes,
  Source = "IDsource",
  Target = "IDtarget",
  Value  = "value",
  NodeID = "name",
  NodeGroup = "Level",
  LinkGroup = "MAG",
  fontSize = 14,
  nodeWidth = 22
)

# -----------------------------
# 2. Apply custom coloring via JS
# -----------------------------
sankey_plot <- htmlwidgets::onRender(
  sankey_plot,
  sprintf(
    '
    function(el, x) {
    
       // Force Helvetica font
      d3.select(el)
        .style("font-family", "Helvetica, Arial, sans-serif");

      // Node color scale (by Level)
      var levelColors = d3.scaleOrdinal()
        .domain([%s])
        .range([%s]);

      // Link color scale (by MAG)
      var magColors = d3.scaleOrdinal()
        .domain([%s])
        .range([%s]);

      // Apply node colors
      d3.select(el).selectAll(".node rect")
        .style("fill", function(d) {
          return levelColors(d.group);
        });

      // Apply link colors
      d3.select(el).selectAll(".link")
        .style("stroke", function(d) {
          return magColors(d.group);
        })
        .style("stroke-opacity", 0.4);
    }
    ',
    level_domain, level_range,
    mag_domain, mag_range
  )
)

# -----------------------------
# 3. Display
# -----------------------------
sankey_plot

# Save as HTML
saveWidget(sankey_plot, "./Sankey/PD_sankey.html", selfcontained = TRUE)

# Horizontal layout: width > height

webshot2::webshot(
  url = "./Sankey/PD_sankey.html",
  file = "./Sankey/PD_sankey.pdf",
  vwidth = 1500,
  vheight = 900,
  cliprect = "viewport"
)
```


## Healthy

```{r sankey-PD}
sankey_Healthy <- AVS_filtered %>%
  filter(Agreement == "Healthy-consistent") %>%
  filter(!is.na(Pathway_description)) %>%
  separate(MAG_ID,
           into = c("Prefix", "MAG", "Gene"),
           sep = "__",
           remove = TRUE) %>%
  mutate(MAG = gsub("^[a-z]__", "", MAG)) %>%
  select(MAG, Gene, KEGG_Module, Pathway_description, Metabolite) %>%
  distinct()

# Save to CSV
write.csv(
  sankey_Healthy,
  file = "./Sankey/Healthy_sankey_table.csv",
  row.names = FALSE,
  quote = FALSE
)
```


```{r sankey-links}
# Define functional hierarchy
func_levels <- c(
  "MAG",
  "Gene",
  "KEGG_Module",
  "Pathway_description",
  "Metabolite"
)

# Build functional connections
Sankey_links <- sankey_Healthy %>%
  select(
    MAG,
    Gene,
    KEGG_Module,
    Pathway_description,
    Metabolite
  ) %>%
  distinct() %>%
  mutate(row_id = row_number()) %>%
  pivot_longer(
    cols = all_of(func_levels), 
    names_to = "level",
    values_to = "node"
  ) %>%
  mutate(level = factor(level, levels = func_levels)) %>%
  arrange(row_id, level) %>%
  group_by(row_id) %>%
  mutate(
    next_node = lead(node),
    MAG_link = first(node[level == "MAG"])
  ) %>%
  ungroup() %>%
  filter(!is.na(next_node)) %>%
  count(MAG_link, node, next_node, name = "value") %>%
  rename(
    source = node,
    target = next_node,
    MAG = MAG_link
  )


```

```{r sankey-nodes}
# Build nodes
Sankey_nodes <- data.frame(
  name = unique(c(Sankey_links$source, Sankey_links$target)),
  stringsAsFactors = FALSE
)

# Assign node groups by hierarchy level
Sankey_nodes <- Sankey_nodes %>%
  mutate(Level = case_when(
    name %in% sankey_Healthy$MAG ~ "MAG",
    name %in% sankey_Healthy$Gene ~ "Gene",
    name %in% sankey_Healthy$KEGG_Module ~ "KEGG_Module",
    name %in% sankey_Healthy$Pathway_description ~ "KEGG_Pathway",
    name %in% sankey_Healthy$Metabolite ~ "Metabolite",
    TRUE ~ "Other"
  ))

# Arrange nodes by hierarchy order
Sankey_nodes <- Sankey_nodes %>%
  mutate(Level = factor(
    Level,
    levels = c("MAG", "Gene", "KEGG_Module", "KEGG_Pathway", "Metabolite")
  )) %>%
  arrange(Level)

# Add numeric IDs after arranging
Sankey_nodes <- Sankey_nodes %>%
  mutate(nodeID = row_number() - 1)

Sankey_links <- Sankey_links %>%
  mutate(
    IDsource = match(source, Sankey_nodes$name) - 1,
    IDtarget = match(target, Sankey_nodes$name) - 1
  )

```

```{r sankey-plot, echo=FALSE, message=FALSE, warning=FALSE}
# Generate palette
# Node levels
level_colors <- c(
  MAG           = pal_npg("nrc")(10)[1],
  Gene          = pal_npg("nrc")(10)[3],
  KEGG_Module   = pal_npg("nrc")(10)[4],
  KEGG_Pathway  = pal_npg("nrc")(10)[2],
  Metabolite    = pal_npg("nrc")(10)[5]
)

# MAGs for link colors
base_palette <- c(
  "#0072B2",  # blue
  "#D55E00",  # vermillion/red
  "#009E73",  # green
  "#CC79A7",  # pink/magenta
  "#F0E442",  # yellow
  "#56B4E9",  # sky blue
  "#E69F00",  # orange
  "#000000"   # black
)

MAGs <- unique(Sankey_links$MAG)
n_mag <- length(MAGs)
mag_palette <- if (n_mag <= length(base_palette)) {
  base_palette[1:n_mag]
} else {
  colorRampPalette(base_palette)(n_mag)
}

# Prepare JS domain/range strings for injection
mag_domain <- paste0('"', MAGs, '"', collapse = ",")
mag_range  <- paste0('"', mag_palette, '"', collapse = ",")

level_domain <- paste0('"', names(level_colors), '"', collapse = ",")
level_range  <- paste0('"', level_colors, '"', collapse = ",")
```

```{r sankey-plot, echo=FALSE, message=FALSE, warning=FALSE}

# -----------------------------
# 1. Create Sankey network
# -----------------------------
sankey_plot <- sankeyNetwork(
  Links = Sankey_links,
  Nodes = Sankey_nodes,
  Source = "IDsource",
  Target = "IDtarget",
  Value  = "value",
  NodeID = "name",
  NodeGroup = "Level",
  LinkGroup = "MAG",
  fontSize = 14,
  nodeWidth = 22
)

# -----------------------------
# 2. Apply custom coloring via JS
# -----------------------------
sankey_plot <- htmlwidgets::onRender(
  sankey_plot,
  sprintf(
    '
    function(el, x) {
    
       // Force Helvetica font
      d3.select(el)
        .style("font-family", "Helvetica, Arial, sans-serif");

      // Node color scale (by Level)
      var levelColors = d3.scaleOrdinal()
        .domain([%s])
        .range([%s]);

      // Link color scale (by MAG)
      var magColors = d3.scaleOrdinal()
        .domain([%s])
        .range([%s]);

      // Apply node colors
      d3.select(el).selectAll(".node rect")
        .style("fill", function(d) {
          return levelColors(d.group);
        });

      // Apply link colors
      d3.select(el).selectAll(".link")
        .style("stroke", function(d) {
          return magColors(d.group);
        })
        .style("stroke-opacity", 0.4);
    }
    ',
    level_domain, level_range,
    mag_domain, mag_range
  )
)

# -----------------------------
# 3. Display
# -----------------------------
sankey_plot

# Save as HTML
saveWidget(sankey_plot, "./Sankey/Healthy_sankey.html", selfcontained = TRUE)

# Horizontal layout: width > height

webshot2::webshot(
  url = "./Sankey/Healthy_sankey.html",
  file = "./Sankey/Healthy_sankey.pdf",
  vwidth = 1500,
  vheight = 300,
  cliprect = "viewport"
)
```

# 3. Enrichment Analysis

## Load differential expression genes per MAG

```{r DEGs}
DEGs <- read.csv("./Data/PDvsH_results.csv", header = TRUE, sep = ",", stringsAsFactors = FALSE) %>%
  mutate(Group = ifelse(regulation == "Upregulated", "PD",
                 ifelse(regulation == "Downregulated", "Healthy", NA)))

```

## Create TERM2GENE

```{r term2gene}

# GO enrichment
term2gene_go <- eggnog %>%
  select(query, GOs) %>%
  filter(
    !is.na(GOs), GOs != "-"
  ) %>%
  separate_rows(GOs, sep = "[,;]") %>%
  mutate(
    GOs = stringr::str_trim(GOs)
  ) %>%
  select(term = GOs, gene = query) %>%
  distinct()

# KEGG KO enrichment
term2gene_kegg <- eggnog %>%
  select(query, KEGG_Pathway) %>%
  filter(
    !is.na(KEGG_Pathway), KEGG_Pathway != "-"
  ) %>%
  separate_rows(KEGG_Pathway, sep = "[,;]") %>%
  mutate(
    KEGG_Pathway = stringr::str_trim(KEGG_Pathway)
  ) %>%
  filter(grepl("^ko", KEGG_Pathway)) %>%
  select(term = KEGG_Pathway, gene = query) %>%
  distinct()

# Check the results
cat("GO terms:", length(unique(term2gene_go$term)), "\n")
cat("GO genes:", length(unique(term2gene_go$gene)), "\n")
cat("KEGG terms:", length(unique(term2gene_kegg$term)), "\n")
cat("KEGG genes:", length(unique(term2gene_kegg$gene)), "\n")
```

## Create TERM2NAME

```{r term2name}


# =====================================================
# GO enrichment
# =====================================================

# GO → Term / Definition

get_go_terms <- function(go_vec) {

  go_vec %>%
    unique() %>%
    na.omit() %>%
    purrr::map_dfr(function(go_id) {

      tibble(
        GOs = go_id,
        GO_Term = tryCatch(
          AnnotationDbi::Term(GO.db::GOTERM[[go_id]]),
          error = function(e) NA_character_
        ),
        GO_Definition = tryCatch(
          AnnotationDbi::Definition(GO.db::GOTERM[[go_id]]),
          error = function(e) NA_character_
        )
      )
    })
}

go_terms <- get_go_terms(term2gene_go$term)

# term2name
term2name_go <- go_terms %>%
  select(term = GOs, name = GO_Term) %>%
  filter(!is.na(name)) %>%
  distinct()

# =====================================================
# KEGG KO enrichment
# =====================================================

# Pathway → Description

get_pathway_descriptions <- function(pathway_vec) {

  pathway_vec %>%
    unique() %>%
    na.omit() %>%
    map_dfr(function(pid_raw) {

      pid <- sub("^path:", "", pid_raw)

      info <- tryCatch(
        keggGet(pid)[[1]],
        error = function(e) NULL
      )

      if (is.null(info) || is.null(info$NAME)) {
        tibble(
          KEGG_Pathway = pid,
          Pathway_description = NA_character_
        )
      } else {
        tibble(
          KEGG_Pathway = pid,
          Pathway_description = paste(info$NAME, collapse = "; ")
        )
      }
    })
}

pathway_desc <- get_pathway_descriptions(term2gene_kegg$term)

# term2name
term2name_kegg <- pathway_desc %>%
  select(term = KEGG_Pathway, name = Pathway_description) %>%
  filter(!is.na(name)) %>%
  distinct()

# Check the results
cat("GO terms:", length(unique(term2name_go$term)), "\n")
cat("GO names:", length(unique(term2name_go$name)), "\n")
cat("KEGG terms:", length(unique(term2name_kegg$term)), "\n")
cat("KEGG names:", length(unique(term2name_kegg$name)), "\n")
```

# GO Terms

```{r comparecluster-go}
# Set up input
MAG_go <- MAG_genus %>%
  left_join(
    eggnog %>% select(query, GOs),
    by = c("gene_id" = "query")
  ) %>%
  filter(!is.na(GOs), GOs != "-") %>%
  separate_rows(GOs, sep = "[,;]") %>%
  mutate(GOs = stringr::str_trim(GOs)) %>%
  inner_join(term2name_go, by = c("GOs" = "term")) %>%
  select(
    gene  = gene_id,
    Group,
    GOs,
    name
  ) %>%
  distinct()
```

```{r comparecluster-go, message=FALSE, warning=FALSE}
# Run compareCluster using Group as condition
compareGO <- compareCluster(
  gene ~ Group,
  data = MAG_go,
  fun = "enricher",
  TERM2GENE = term2gene_go,
  TERM2NAME = term2name_go,
  universe = eggnog$query,
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.05
)

go_res <- as.data.frame(compareGO)
```

## Dotplot of GO enrichment

```{r comparecluster-go, message=FALSE, warning=FALSE}

# Dotplot of GO enrichment per condition (PD vs Healthy) using MAG genes
MAG_goplot <- dotplot(compareGO, x = "Group", showCategory = 5) +
  scale_color_gradientn(
    colours = c("#f7ca64", "#46bac2", "#371ea3"),
    guide = guide_colorbar(reverse = TRUE)
  ) +
  guides(size = guide_legend(override.aes = list(shape = 16))) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y = element_text(size = 12, color = "black"),
    axis.text.x = element_text(size = 14, angle = 45, hjust = 1, color = "black"),
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  ) +
  xlab(NULL) +
  ylab(NULL)

MAG_goplot

# Save 
ggsave("./Enrichment/MAG_go.pdf", plot = MAG_goplot, dpi = 600, width = 8, height = 4)
write.csv(go_res, "./Enrichment/MAG_go_results.csv", row.names = FALSE)

```

# KEGG

```{r comparecluster-kegg}
# Set up input
MAG_kegg <- MAG_genus %>%
  left_join(
    eggnog %>% select(query, KEGG_Pathway),
    by = c("gene_id" = "query")
  ) %>%
  filter(!is.na(KEGG_Pathway), KEGG_Pathway != "-") %>%
  separate_rows(KEGG_Pathway, sep = "[,;]") %>%
  mutate(KEGG_Pathway = stringr::str_trim(KEGG_Pathway)) %>%
  inner_join(term2name_kegg, by = c("KEGG_Pathway" = "term")) %>%
  filter(grepl("^ko", KEGG_Pathway)) %>%
  select(
    gene  = gene_id,
    Group,
    KEGG_Pathway,
    name
  ) %>%
  distinct()
```

```{r comparecluster-kegg, message=FALSE, warning=FALSE}

# KEGG functional enrichment across MAG Groups
compareKEGG <- compareCluster(
  gene ~ Group,
  data = MAG_kegg,
  fun = "enricher",
  TERM2GENE = term2gene_kegg,
  TERM2NAME = term2name_kegg,
  universe = eggnog$query,
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.05
)

kegg_res <- as.data.frame(compareKEGG)
```

# Dotplot of KEGG enrichment

```{r comparecluster-kegg, message=FALSE, warning=FALSE}

MAG_keggplot <- dotplot(compareKEGG, 
                        x = "Group", 
                        showCategory = 5) +
  scale_colour_distiller(palette = "YlGnBu", direction = 1) +
  guides(size = guide_legend(override.aes = list(shape = 16))) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y = element_text(size = 12, color = "black"),
    axis.text.x = element_text(size = 14, angle = 45, hjust = 1, color = "black"),
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  ) +
  xlab(NULL) +
  ylab(NULL)

MAG_keggplot

# Save 
ggsave("./Enrichment/MAG_kegg.pdf", plot = MAG_keggplot, dpi = 600, width = 8, height = 4)
write.csv(kegg_res, "./Enrichment/MAG_kegg_results.csv", row.names = FALSE)
```

# Session information

```{r, session-information}
sessionInfo()
```
