---
title: "Metatranscriptomic Parkinson’s disease (DeNoPa) study"
author: Alexis Felipe Rojas Cruz
output:
  html_document:
    toc: true
    toc_float: true
    theme: cerulean  
    highlight: tango
---

The aim of this study was to characterize the gut microbiome’s transcriptional activity in Parkinson’s disease (PD) by performing metatranscriptomic profiling of fecal samples from participants in the DeNoPa (De Novo Parkinson’s) cohort. Through differential gene expression and functional annotation, this analysis seeks to identify microbial pathways and functional signatures associated with PD pathophysiology.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.keep = 'all'
)
```


```{r include=FALSE, comment = ""}
# Define packages and their sources
bioc_packages <- c("DESeq2", "tximport", "BiocParallel", "apeglm", "PCAtools")
cran_packages <- c("RColorBrewer", "statmod", "tidyverse", "ggplot2", "ggrepel", "pheatmap", "readr")

# Install Bioconductor packages
missing_bioc <- bioc_packages[!bioc_packages %in% rownames(installed.packages())]
if (length(missing_bioc) > 0) {
  if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
  }
  BiocManager::install(missing_bioc)
}

# Install CRAN packages
missing_cran <- cran_packages[!cran_packages %in% rownames(installed.packages())]
if (length(missing_cran) > 0) {
  install.packages(missing_cran)
}

# Load all libraries
library(DESeq2)
library(tximport)
library(BiocParallel)
library(apeglm)
library(PCAtools)
library(tidyverse)
library(pheatmap)
library(ggrepel)
library(RColorBrewer)
library(statmod)
library(readr)

```

# Set up directory and find salmon files


```{r find_files}
# Define directory where Salmon quantifications are stored
salmon_dir <- "~/Documents/Parkinson/Metatranscriptomics/Salmon_out/"

# List all quant.sf files
files <- list.files(path = salmon_dir, pattern = "quant.sf", recursive = TRUE, full.names = TRUE)

# Check if any files were found
if (length(files) == 0) {
  stop("No quant.sf files found in ", salmon_dir, ". Please check the path.")
}

# Extract sample names from directory names
sample_names <- basename(dirname(files))
names(files) <- sample_names

cat("Found", length(files), "samples:\n")
print(sample_names)
```

# Load Transcript-to-Gene mapping

```{r load_tx2gene}

tx2gene_raw <- read.delim("Salmon_out/MAGxTranscripts_modified.tsv", header = FALSE, stringsAsFactors = FALSE)

# keep only the first two:

tx2gene <- tx2gene_raw %>%
  dplyr::select(V1, V2) %>%
  dplyr::rename(transcript_id = V1, gene_id = V2)

# Remove duplicates and missing values just in case
tx2gene <- tx2gene_raw %>%
  dplyr::select(V1, V2) %>%
  dplyr::rename(transcript_id = V1, gene_id = V2)

cat("tx2gene mapping contains", nrow(tx2gene), "unique transcript-gene pairs.\n")
print(tx2gene)
```

# Import data with tximport

```{r tximport_data}
# Import data with tximport
txi <- tximport(
  files = files,
  type = "salmon",
  tx2gene = tx2gene,
  countsFromAbundance = "lengthScaledTPM"
)

cat("Tximport completed successfully!\n")

# Save RD
saveRDS(txi, file = "~/Documents/Parkinson/Metatranscriptomics/tximport_results.rds")
```

# Check import results

```{r check_results}
# Read the saved tximport object
txi <- readRDS("~/Documents/Parkinson/Metatranscriptomics/Expression/tximport/tximport_results.rds")

# Check import results
cat("Summary of imported data:\n")
cat("Counts matrix dimensions:", dim(txi$counts), "\n")
cat("Sample names:", paste(colnames(txi$counts), collapse = ", "), "\n")
cat("Counts range:", round(range(txi$counts), 2), "\n")
cat("Total counts per sample:\n")
print(colSums(txi$counts))
```
# Annotation

```{r annotation}
# Read the merged taxonomy file of MAGs
taxon <- read.delim("Annotation/MAG_taxonomy.tsv", sep = "\t", header = TRUE)

# Read the EggNOG merged annotation file
eggnog <- read.delim("Annotation/MAG_annotations.tsv", sep = "\t", header = TRUE)

# Merge EggNOG functional annotations with taxonomic classification
MAG_anno <- eggnog %>%
  tidyr::separate(query, into = c("user_genome", "contig_name", "gene_number"),
                  sep = "\\|", remove = FALSE) %>%
  select(-contig_name, -gene_number) %>%
  left_join(
    taxon %>% select(user_genome, classification),
    by = "user_genome"
  ) %>%
  relocate(user_genome, classification, .after = query) %>%
  tidyr::separate(
    classification,
    into = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"),
    sep = ";", fill = "right", remove = FALSE
  ) %>%
  mutate(MAG_gene = paste(Species, Preferred_name, sep = "__")) %>%
  relocate(MAG_gene, .after = query)

# Save the final annotation file
write.table(MAG_anno, file = "./Annotation/Final_annotation.tsv", sep = "\t", quote = FALSE, row.names = FALSE)
```

# Visual check of sample Statistics


```{r visual-check}
# Summarize counts
sample_stats <- data.frame(
  Sample = colnames(txi$counts),
  Total_Counts = colSums(txi$counts),
  Nonzero_Genes = colSums(txi$counts > 0),
  Mean_Counts = colMeans(txi$counts)
)

# Plot library sizes
ggplot(sample_stats, 
       aes(x = Sample, 
           y = Total_Counts, 
           fill = ifelse(grepl("CT", Sample, ignore.case = TRUE), "Healthy", "PD"))) +
  geom_bar(stat = "identity") +
  scale_fill_manual(
    values = c("Healthy" = "steelblue", "PD" = "darkorange"), 
    name = "Group"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  ) +
  labs(
    y = "Total Counts", 
    x = "Samples"
  )

``` 

#  Identify low-count samples

```{r low-count samples}
# Filter samples < 100K counts
low_samples <- sample_stats %>%
  filter(Total_Counts < 1e5) %>%
  pull(Sample)
print(low_samples)

# Filter txi object accordingly
txi_filtered <- txi
keep_samples <- setdiff(colnames(txi$counts), low_samples)

txi_filtered$counts <- txi$counts[, keep_samples]
txi_filtered$abundance <- txi$abundance[, keep_samples] 
txi_filtered$length <- txi$length[, keep_samples]

```

# Create sample metadata

```{r sample metadata}
# Create colData by keeping only the filtered samples
colData <- sample_stats %>%
  rownames_to_column(var = "rowname") %>%   
  mutate(Group = case_when(
    str_starts(Sample, "CT_") ~ "Healthy",    
    str_starts(Sample, "PD_") ~ "PD",
    TRUE ~ "Unknown"
  )) %>%
  filter(!Sample %in% low_samples) %>%
  select(Sample, Group) %>%
  mutate(Group = factor(Group)) %>%
  column_to_rownames("Sample")

# Verify
all(rownames(colData) == colnames(txi_filtered$counts))

```

# Create DESeq2 object

```{r DESeq2-object}
# Create DESeq2 dataset
meta_dds <- DESeqDataSetFromTximport(
  txi_filtered,
  colData = colData,
  design = ~ Group)

meta_dds

head(rownames(meta_dds))

```

# Filtering low count genes

```{r low-count genes}

# Pre-filter: remove genes with very low counts across all samples
keep <- rowSums(counts(meta_dds)) >= 10
meta_dds <- meta_dds[keep,]

cat("After pre-filtering:\n")
cat("Original genes:", nrow(txi_filtered$counts), "\n")
cat("Genes after filtering:", nrow(meta_dds), "\n")
cat("Percentage retained:", round(nrow(meta_dds)/nrow(txi_filtered$counts) * 100, 2), "%\n")
```
# Run DESeq2 Analysis

```{r DESeq2}
# Run DESeq2 analysis
meta_dds <- DESeq(meta_dds)

cat("Results names:", resultsNames(meta_dds), "\n")

```

# Estimating the dispersion

Plotting the dispersion estimates is a useful diagnostic. The dispersion plot below is typical, with the final estimates shrunk from the gene-wise estimates towards the fitted estimates.

```{r, Disp}
# Check dispersion estimates
plotDispEsts(meta_dds)
```

# Variance Stabilizing Transformation

```{r VST}
# VST normalization
meta_vst <- vst(meta_dds, blind = FALSE)

# Extract vst-normalized counts
vst_counts <- assay(meta_vst)

# Save as RDS for reproducibility
saveRDS(vst_counts, "VST_normalization/meta_vst_counts.rds")
```

# Quality control

## PCA

```{r PCA, warning=FALSE}
# Ensure Group column exists and is a factor
colData(meta_vst)$Group <- factor(colData(meta_vst)$Group)

# Perform PCA using vst-normalized counts
pca_res <- pca(vst_counts,
               metadata = colData(meta_vst))

# Calculate explained variance
explained_var_percent <- round(100 * (pca_res$variance / sum(pca_res$variance)), 2)

# Label axes with explained variance
xlab <- paste0("PC1 (", explained_var_percent[1], "%)")
ylab <- paste0("PC2 (", explained_var_percent[2], "%)")

# Define shapes and custom colors
points <- c(21, 22)
colors <- c("Healthy" = "steelblue",  
            "PD"      = "darkorange")   

# PCA biplot
biplot(pca_res,
       colby = "Group",                      
       colkey = colors,
       shape = "Group",                      
       shapekey = setNames(points, levels(colData(meta_vst)$Group)),
       pointSize = 3,
       legendPosition = 'right',
       title = 'PCA of VST-normalized Expression',
       xlab = xlab,
       ylab = ylab)


```


## Distance heatmap

```{r, heatmap}
# Compute pairwise sample distances using vst-normalized counts
sampleDists <- dist(t(assay(meta_vst)))

# Convert to matrix for heatmap
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- colnames(meta_vst)
colnames(sampleDistMatrix) <- colnames(meta_vst)

# Annotation data for columns (Group info)
annotation_col <- data.frame(
  Group = colData(meta_vst)$Group,
  row.names = colnames(meta_vst)
)

# Custom color palette for groups
ann_colors <- list(
  Group = c("Healthy" = "steelblue",
            "PD" = "darkorange")
)

# Plot heatmap of sample-to-sample distances
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         color = colorRampPalette(rev(brewer.pal(9, "Blues")))(255),
         annotation_col = annotation_col,
         annotation_colors = ann_colors,
         main = "Heatmap of Sample Distances (VST-normalized)",
         fontsize_row = 8,
         fontsize_col = 8)

```


# Functions

```{r formatting}
# Format and merge DESeq2 results with EggNOG + taxonomy annotations
format_DESeq2_results <- function(DESeq2_results, MAG_anno) {
  DESeq2_results_fmt <- DESeq2_results %>%
    as.data.frame() %>%
    rownames_to_column(var = "gene_id") %>% 
    select(
      gene_id,
      baseMean,
      log2FoldChange,
      lfcSE,
      pvalue,
      padj
    ) %>%
    filter(!is.na(pvalue), !is.na(padj)) %>%
    arrange(padj, pvalue)
  
  MAG_anno_fmt <- MAG_anno %>%
    rename(gene_id = query)
  
  merged_results <- DESeq2_results_fmt %>%
    left_join(
      MAG_anno_fmt %>%
        select(
          gene_id,
          MAG_gene,
          classification,
          Kingdom,
          Phylum,
          Class,
          Order,
          Family,
          Genus,
          Species,
          Preferred_name,
          Description
        ),
      by = "gene_id"
    )
  
  merged_results <- merged_results %>%
    select(
      gene_id,
      MAG_gene,
      classification,
      Kingdom,
      Phylum,
      Class,
      Order,
      Family,
      Genus,
      Species,
      Preferred_name,
      Description,
      baseMean,
      log2FoldChange,
      lfcSE,
      pvalue,
      padj
    )
  
  return(merged_results)
}

```

```{r summarize}
# Summarize number of DEGs (up/down)
summarize_pretty_deg <- function(pretty_deg, lfc_cutoff = 1, padj_cutoff = 0.05) {

  down <- pretty_deg %>%
    filter(padj <= padj_cutoff & log2FoldChange <= -1.0) %>%
    nrow()

  down_fc <- pretty_deg %>%
    filter(padj <= padj_cutoff & log2FoldChange <= (-1 * lfc_cutoff)) %>%
    nrow()

  up <- pretty_deg %>%
    filter(padj <= padj_cutoff & log2FoldChange >= 1.0) %>%
    nrow()

  up_fc <- pretty_deg %>%
    filter(padj <= padj_cutoff & log2FoldChange >= lfc_cutoff) %>%
    nrow()

  data.frame(
    Change = c('Up', paste0('Up - min ', sprintf(fmt="%.2f", lfc_cutoff)), 
               'Down', paste0('Down - min ', sprintf(fmt="%.2f", lfc_cutoff))),
    n = c(up, up_fc, down, down_fc)
  ) %>%
    return()
}
```

```{r extract}
# Extract significantly up/downregulated genes
get_DEGs <- function(deg_df, padj_cutoff = 0.05, lfc_cutoff = 1) {
  deg_df %>%
    mutate(regulation = case_when(
      padj < padj_cutoff & log2FoldChange >= lfc_cutoff ~ "Upregulated",
      padj < padj_cutoff & log2FoldChange <= -lfc_cutoff ~ "Downregulated",
      TRUE ~ "Not"
    )) %>%
    filter(regulation != "Not") %>%
    arrange(desc(log2FoldChange)) %>%
    return()
}

```


# Differential expression

Now we will find differentially expressed genes (DEGs) among combinations


## Contrasts

Use Treatment - Control if you want to assess whether the treatment increases the expression relative to control.
Use Control - Treatment if you want to assess whether the control increases the expression relative to the treatment.

For this study we will perform this comparison:


- `PD vs Healthy`

# PD vs Healthy

```{r PDvsHealthy-DESeq2, message=FALSE, warning=FALSE}
# Differential expression analysis: PD vs Healthy
res_PDvsHealthy <- results(meta_dds, 
                           contrast = c("Group", "PD", "Healthy"), 
                           pAdjustMethod = "BH")

# Summary of results
summary(res_PDvsHealthy, alpha = 0.05)

```

## Shrink log2 fold changes with apeglm

```{r PDvsHealthy-apeglm, message=FALSE, warning=FALSE}

res_PDvsHealthy_sh <- lfcShrink(meta_dds,
                                    coef = "Group_PD_vs_Healthy",
                                    type = "apeglm")

summary(res_PDvsHealthy_sh, alpha = 0.05)
```


## MA Plot

```{r PDvsHealthy-MAplot, fig.height=5, fig.width=7}
par(mfrow = c(1, 2))

# Unshrunken MA plot
plotMA(res_PDvsHealthy,
       main = "Unshrunken: PD vs Healthy",
       ylim = c(-5, 5))
abline(h = c(-1.5, 1.5), col = "dodgerblue", lwd = 2)

# Shrunken (apeglm) MA plot
plotMA(res_PDvsHealthy_sh,
       main = "Shrunken: PD vs Healthy",
       ylim = c(-5, 5))
abline(h = c(-1.5, 1.5), col = "dodgerblue", lwd = 2)

```

## Formatting

```{r PDvsHealthy-format}
# Format DESeq2 results unshrunken
PDvsHealthy_res <- format_DESeq2_results(res_PDvsHealthy, MAG_anno)
head(PDvsHealthy_res)

```


```{r PDvsHealthy-format}
# Format DESeq2 results shrunken
PDvsHealthy_sh <- format_DESeq2_results(res_PDvsHealthy_sh, MAG_anno)
head(PDvsHealthy_sh)

```

## Summarize

```{r PDvsHealthy-summary}
# Summarize up/down-regulated gene counts
PDvsHealthy_summary <- summarize_pretty_deg(PDvsHealthy_sh,
                                            lfc_cutoff = 1.5,
                                            padj_cutoff = 0.05)
PDvsHealthy_summary
```

## Get DEGs

```{r PDvsHealthy-DEGs}
# Filter DEGs: |log2FC| ≥ 1.5 and padj < 0.05
PDvsHealthy_DEGs <- get_DEGs(PDvsHealthy_sh,
                             padj_cutoff = 0.05,
                             lfc_cutoff = 0)
head(PDvsHealthy_DEGs)

```

# Save data

```{r save}
# Save the differential expressed genes
write.csv(PDvsHealthy_DEGs, file = "./PDvsHealthy/All_PDvsH_results2.csv",
            quote = TRUE, row.names = FALSE)

```


# Volcano plot

```{r volcano_PDvsHealthy, fig.width=10, fig.height=8, message=FALSE, warning=FALSE}

# Convert DESeq2 results to a data frame
volcano_data <- as.data.frame(PDvsHealthy_res) %>%
  mutate(
    log10_padj = -log10(padj),
    significance = case_when(
      padj < 0.05 & log2FoldChange > 0 ~ "PD Enriched",
      padj < 0.05 & log2FoldChange < 0 ~ "Healthy Enriched",
      TRUE ~ "Not Significant"
    ),
    label = ifelse(padj < 0.05 & abs(log2FoldChange) > 1.5, gene_id, NA)
  )

# Define colors (from your original code)
color_palette <- c(
  "PD Enriched" = "#8c6bb1",       # Purple
  "Healthy Enriched" = "#fe9929",  # Orange
  "Not Significant" = "gray70"
)

# Create volcano plot
volcano_plot <- ggplot(volcano_data, aes(x = log2FoldChange, y = log10_padj)) +
  geom_point(aes(color = significance, fill = significance),
             alpha = 0.7, size = 2, shape = 21) +
  scale_color_manual(values = color_palette) +
  scale_fill_manual(values = color_palette) +

  # Add thresholds
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed", color = "red", alpha = 0.5) +
  geom_vline(xintercept = 0,
             linetype = "dashed", color = "black", alpha = 0.5) +
  geom_vline(xintercept = c(-1.5, 1.5),
             linetype = "dotted", color = "blue", alpha = 0.3) +

  # Theme and labels
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "right",
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    text = element_text(size = 12),
    plot.title = element_text(face = "bold", size = 14)
  ) +
  labs(
    x = "Log2 Fold Change (PD vs Healthy)",
    y = "-log10(FDR-adjusted p-value)",
    color = "Significance",
    fill = "Significance"
  ) +
  coord_cartesian(clip = "off")

# Print volcano plot
print(volcano_plot)

# Save the plot
ggsave("./PDvsHealthy/Volcano_PDvsHealthy.pdf",
       plot = volcano_plot,
       width = 10, height = 8, dpi = 300)

```

## Heatmap 

```{r heatmap}
# Parameters
padj_cutoff <- 0.05
lfc_cutoff <- 1.5
n_top <- 50
condition_column <- "Group"

# Filter significant DEGs
deg_df <- as.data.frame(PDvsHealthy_sh) %>%
  filter(padj < padj_cutoff & abs(log2FoldChange) > lfc_cutoff)

if (nrow(deg_df) == 0) stop("No significant DEGs found with current thresholds.")

# Select exactly 25 up and 25 down DEGs
up_genes <- deg_df %>% 
  filter(log2FoldChange > 0) %>% 
  arrange(padj) %>% 
  slice_head(n = n_top/2) %>%
  pull(gene_id)

down_genes <- deg_df %>% 
  filter(log2FoldChange < 0) %>% 
  arrange(padj) %>% 
  slice_head(n = n_top/2) %>%
  pull(gene_id)

top_genes <- c(up_genes, down_genes)

cat("Selected", length(up_genes), "up-regulated and", length(down_genes), "down-regulated genes\n")

# Use variance-stabilized transformed data (VST)
vst_mat <- assay(meta_vst)[top_genes, , drop = FALSE]

# Z-score normalization by gene
zscore_data <- t(scale(t(vst_mat)))

# Remove constant rows (if any)
zscore_data <- zscore_data[complete.cases(zscore_data), ]

# Replace rownames with readable MAG_gene identifiers
gene_map <- deg_df %>% select(gene_id, MAG_gene) %>% distinct()
rownames(zscore_data) <- gene_map$MAG_gene[match(rownames(zscore_data), gene_map$gene_id)]

# Sample annotation (condition labels)
annotation_col <- data.frame(
  Group = colData(meta_dds)[[condition_column]],
  row.names = colnames(meta_dds)
)

# Define color palettes
group_colors <- list(
  Group = c(Healthy = "#fe9929", PD = "#8c6bb1")
)

heatmap_colors <- colorRampPalette(brewer.pal(11, "RdYlBu"))(100)

# Plot heatmap
heatmap <- pheatmap(
  zscore_data,
  color = heatmap_colors,
  annotation_col = annotation_col,
  annotation_colors = group_colors,
  cluster_rows = FALSE,
  clustering_method = "ward.D2",  
  clustering_distance_rows = "correlation", 
  clustering_distance_cols = "correlation",
  show_rownames = TRUE,
  show_colnames = ncol(zscore_data) <= 30,
  fontsize_row = 7,
  fontsize_col = 8,
  border_color = NA,
  scale = "none"
)

heatmap

# Save the plot
ggsave("./PDvsHealthy/Heatmap_PDvsHealthy.pdf",
       plot = heatmap,
       width = 10, height = 8, dpi = 300)

```


## Session info

```{r session_information}
Sys.time()
getwd()
  
sessionInfo()
```